---
title: "ForestForTrees: Cross-Biome Benchmark and Performance Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ForestForTrees: Cross-Biome Benchmark and Performance Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(ForestForTrees)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(sf)
options(mc.cores = 4)
```

## Overview

This vignette serves two primary purposes:

**Empirical Validation:** We compare the package's estimates of the power-law exponent (α) and total stem abundance (N_tot) against field-measured data from 20 National Ecological Observatory Network (NEON) sites spanning subtropical to alpine forest types.

**Performance Benchmarking:** We record execution times for the key computational steps — breakpoint detection, data truncation, and Bayesian inference — to demonstrate the workflow's scalability and help users assess the practicality of large-scale deployment.

Note that the same NEON dataset was used in Eichenwald et al. (2025b) to validate the underlying mathematical framework. Its role here is distinct: rather than demonstrating that the *theory* recovers field-measured values, this vignette stress-tests the *software* and confirms that the automated pipeline converges reliably across radically different forest types without manual reconfiguration.

---

## Multi-Site Alpha Estimation

We define the set of sites to process. Note that ABBY (Abby Road, Washington) is included in the initial run but excluded from the validation plots. ABBY was excluded because the Bayesian R² for that site was below 0.6, indicating that the fitted scaling distribution was a poor match to the observed size-abundance data — likely reflecting the site's highly disturbed, structurally complex stand history. The R² filter applied in the plotting steps below removes ABBY automatically.

```{r}
sitevector <- c("ABBY", "BART", "BLAN", "BONA", "CLBJ", "DELA", "GUAN",
                "HARV", "JERC", "LENO", "MLBS", "OSBS",
                "RMNP", "SCBI", "SERC", "SJER", "SOAP", "TALL", "TEAK",
                "WREF")

results_model   <- list()
results_summary <- data.frame()
```

We iterate over each site, performing breakpoint detection, filtering, and alpha estimation, recording runtime at each step.

```{r warning=FALSE, message=FALSE}
results_model   <- list()
results_summary <- data.frame()
runtime_table   <- data.frame(
  siteID               = character(),
  potential_break_time = numeric(),
  truncate_filter_time = numeric(),
  fit_alpha_model_time = numeric()
)

for (site in sitevector) {

  site_data <- neon_data %>%
    filter(siteID == site & dbh <= 50) %>%
    data.frame()

  if (nrow(site_data %>% filter(dbh > 10)) < 25) next

  # Step 1: Identify potential breakpoint
  t1 <- system.time({
    breakpoint_kde <- potential_break(
      data        = site_data,
      n_bootstrap = 1000,
      trim_max    = 50
    )
  })

  # Step 2: Filter and truncate data based on breakpoint
  t2 <- system.time({
    filtered_data <- truncate_filter(
      breakpoint_kde_results = breakpoint_kde,
      min_size               = 10
    )
  })

  # Step 3: Fit Bayesian alpha model
  site_priors <- alpha_priors %>% filter(siteID == site)

  LAI_value <- Leaf_area_index %>%
    rename(siteID = site) %>%
    filter(siteID == site) %>%
    pull(Leaf_area_index)

  t3 <- system.time({
    fit <- fit_alpha_model(
      filtered_data = filtered_data,
      LAI           = LAI_value,
      prior_mean    = site_priors$prior_mean,
      prior_sd      = site_priors$prior_sd,
      iter          = 9000,
      warmup        = 6000,
      chains        = 4,
      cores         = 4
    )
  })

  results_model[[site]] <- fit
  results_summary <- rbind(
    results_summary,
    fit$posterior_summary %>% mutate(siteID = site)
  )

  runtime_table <- rbind(
    runtime_table,
    data.frame(
      siteID               = site,
      potential_break_time = t1["elapsed"],
      truncate_filter_time = t2["elapsed"],
      fit_alpha_model_time = t3["elapsed"]
    )
  )
}

# Summary statistics
runtime_summary <- runtime_table %>%
  summarise(
    mean_break    = mean(potential_break_time),
    mean_truncate = mean(truncate_filter_time),
    mean_fit      = mean(fit_alpha_model_time),
    median_break  = median(potential_break_time),
    median_fit    = median(fit_alpha_model_time),
    max_fit       = max(fit_alpha_model_time),
    min_fit       = min(fit_alpha_model_time)
  )

runtime_summary
```

---

## Visualizing Alpha Estimates

We compare estimated alpha values to field-measured values. Sites with Bayesian R² < 0.6 are excluded, as low R² indicates a poor fit of the scaling distribution to the observed data. In our analysis, ABBY failed this threshold due to its highly disturbed stand structure, and is therefore excluded from the validation.

```{r fig.height=5, fig.width=6}
data("neon_truemeasuredalpha")

ggplot(
  results_summary %>%
    inner_join(neon_truemeasuredalpha, by = "siteID") %>%
    filter(R2_kernel > 0.6),
  aes(x = alpha, y = mean, label = siteID)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(
    x = "Field-Measured Alpha",
    y = "ForestForTrees Estimated Alpha"
  ) +
  theme_minimal() +
  geom_smooth(method = "lm")
```

---

## Estimation of Total Stem Abundance (N_tot)

Following estimation of α for each site, we estimate total stem abundance (N_tot). This step propagates uncertainty from the fitted alpha model into posterior estimates of total tree density.

```{r}
results_abundance_model   <- list()
results_abundance_summary <- data.frame()

abundance_runtime <- data.frame(
  siteID      = character(),
  elapsed_sec = numeric(),
  stringsAsFactors = FALSE
)

for (site in names(results_model)) {

  message(paste("Estimating total abundance for site:", site))

  alpha_fit <- results_model[[site]]

  t_start <- Sys.time()

  trees <- estimate_total_trees(
    alpha_model_output  = alpha_fit,
    N_tot_prior_mean    = 1250,
    N_tot_prior_sd      = 625,
    warmup              = 6000,
    iter                = 9000,
    siteID              = site
  )

  t_end <- Sys.time()

  results_abundance_model[[site]] <- trees

  if (!is.null(trees$posterior_summary)) {
    results_abundance_summary <- rbind(
      results_abundance_summary,
      trees$posterior_summary %>% mutate(siteID = site)
    )
  }

  abundance_runtime <- rbind(
    abundance_runtime,
    data.frame(
      siteID      = site,
      elapsed_sec = as.numeric(difftime(t_end, t_start, units = "secs"))
    )
  )
}

abundance_runtime

abundance_runtime_summary <- abundance_runtime %>%
  summarise(
    n_sites    = n(),
    mean_sec   = mean(elapsed_sec),
    median_sec = median(elapsed_sec),
    sd_sec     = sd(elapsed_sec),
    min_sec    = min(elapsed_sec),
    max_sec    = max(elapsed_sec)
  )

abundance_runtime_summary
```

---

## Visualizing Abundance Estimates

We compare estimated abundance values to field-measured values across all sites that successfully ran.

```{r fig.height=5, fig.width=6}
abundancedata <- data.frame(
  siteID = c(
    "ABBY","BART","BLAN","BONA","CLBJ","DEJU","DELA","DSNY","GUAN",
    "HARV","HEAL","JERC","KONZ","LAJA","LENO","MLBS","MOAB","NIWO",
    "OSBS","RMNP","SCBI","SERC","SJER","SOAP","SRER","TALL","TEAK",
    "WREF","YELL"
  ),
  abundance = c(
    633,1965,251,1576,629,984,1269,48,586,
    2111,47,294,40,8,740,1176,42,1042,
    869,1307,741,1055,110,1044,8,1084,1033,
    1225,452
  )
)

ggplot(
  results_abundance_summary %>%
    inner_join(abundancedata, by = "siteID"),
  aes(x = abundance, y = mean, label = siteID)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(
    x = "Field-Measured Abundance",
    y = "ForestForTrees Estimated Abundance"
  ) +
  theme_minimal() +
  geom_smooth(method = "lm")
```

---

## Notes

- The workflow uses **bootstrap KDEs** to identify size distribution peaks and potential breakpoints.
- Filtering ensures the model fits to the **upper tail**, where Pareto behavior is expected.
- LAI and breakpoint corrections account for **reduced detectability of small trees** in remote sensing data.
- Sites with fewer than 25 trees after filtering are skipped to ensure robust estimation.
- ABBY is excluded from validation plots due to low Bayesian R² (poor distributional fit), violating the Pareto scaling assumption.
