---
title: "Estimating Power-Law Exponents from Remote Sensing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Alpha with ScalingFromSkypackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ScalingFromSkypackage)
library(dplyr)
library(ggplot2)
library(rstan)
library(sf)
library(microbenchmark)
options(mc.cores = 4)
```

## Overview

This vignette serves two primary purposes:

Empirical Validation: We compare the package's estimates of the power-law exponent ($\alpha$) and total stem abundance ($N_{tot}$) against field-measured data from the National Ecological Observatory Network (NEON).

Performance Benchmarking: We track execution times for the key computational steps—breakpoint detection, data truncation, and Bayesian inference—to demonstrate the workflow's scalability across multiple sites.

---

## Multi-Site Alpha Estimation

We first define the set of sites to process:

```{r}
sitevector <- c("ABBY", "BART", "BLAN", "BONA", "CLBJ", "DELA", "GUAN",
                "HARV", "JERC", "LENO", "MLBS", "OSBS",
                "RMNP", "SCBI", "SERC", "SJER", "SOAP", "TALL", "TEAK",
                "WREF")

results_model <- list()
results_summary <- data.frame()
```

We iterate over each site, performing breakpoint detection, filtering, and alpha estimation.

```{r, warning=FALSE, message=FALSE}

# Initialize outputs
results_model <- list()
results_summary <- data.frame()
runtime_table <- data.frame(
  siteID = character(),
  potential_break_time = numeric(),
  truncate_filter_time = numeric(),
  fit_alpha_model_time = numeric()
)

for(site in sitevector){

  site_data <- neon_data %>%
    filter(siteID == site & dbh <= 50) %>%
    data.frame()

  # Skip sites with insufficient data
  if(nrow(site_data %>% filter(dbh > 10)) < 25) next

  # Step 1: Identify potential breakpoint
  t1 <- system.time({
    breakpoint_kde <- potential_break(
      data = site_data,
      n_bootstrap = 1000,
      trim_max = 50
    )
  })
  
  # Step 2: Filter and truncate data based on breakpoint
  t2 <- system.time({
    filtered_data <- truncate_filter(
      breakpoint_kde_results = breakpoint_kde,
      min_size = 10
    )
  })

  # Step 3: Fit Bayesian alpha model
  site_priors <- alpha_priors %>% filter(siteID == site)
  
  LAI_value <- Leaf_area_index %>%
    rename(siteID = site) %>%
    filter(siteID == site) %>%
    pull(Leaf_area_index)

  t3 <- system.time({
    fit <- fit_alpha_model(
      filtered_data= filtered_data,
      LAI = LAI_value,
      prior_mean = site_priors$prior_mean,
      prior_sd = site_priors$prior_sd,
      iter = 9000,
      warmup = 6000,
      chains = 4,
      cores = 4
    )
  })

  # Store results
  results_model[[site]] <- fit
  results_summary <- rbind(
  results_summary,
  fit$posterior_summary %>%
  mutate(siteID = site)
  )
  # Record runtimes
  runtime_table <- rbind(
    runtime_table,
    data.frame(
      siteID = site,
      potential_break_time = t1["elapsed"],
      truncate_filter_time = t2["elapsed"],
      fit_alpha_model_time = t3["elapsed"]
    )
  )
}

# Optionally, calculate summary statistics
runtime_summary <- runtime_table %>%
  summarise(
    mean_break = mean(potential_break_time),
    mean_truncate = mean(truncate_filter_time),
    mean_fit = mean(fit_alpha_model_time),
    median_break = median(potential_break_time),
    median_truncate = median(truncate_filter_time),
    median_fit = median(fit_alpha_model_time),
    max_fit = max(fit_alpha_model_time),
    min_fit =min(fit_alpha_model_time)
  )

runtime_summary
```

---

## Visualizing Alpha Estimates

We can quickly compare estimated alpha values to previously reported values for validation. The 1:1 line in red would represent a "perfect" replication of measured data by our interpolation. Notice that we add a filter to remove sites with low Bayesian R2 values, as such values would suggest that the fit of the Bayesian distribution to find ($\alpha$) was poor for that site. In this case, the site "ABBY" was removed from the analysis:

```{r}
data("neon_truemeasuredalpha")

ggplot(results_summary%>%
         inner_join(neon_truemeasuredalpha)%>%
         filter(R2_kernel > 0.6), aes(x = alpha, y = mean, label = siteID)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(nudge_y = 0.1) +
  labs(x = "Reported Alpha", y = "ScalingFromSky Estimated Alpha") +
  theme_minimal() +
  geom_smooth(method = "lm")
```

---

## Notes

- The workflow uses **bootstrap KDEs** to identify size distribution peaks and potential breakpoints.
- Filtering ensures the model fits to the **upper tail**, where Pareto behavior is expected.
- LAI and breakpoint corrections account for **reduced detectability of small trees** in remote sensing data.
- Sites with fewer than 25 trees after filtering are skipped to ensure robust estimation.



## Estimation of Total Stem Abundance (N_tot)


Following estimation of the size–abundance scaling parameter (alpha) for each site, the next stage in the ScalingFromSky workflow is estimation of **total stem abundance (N_tot)**. This step scales the relative size–abundance information inferred from remote sensing to an absolute estimate of tree density at each site.


This is implemented using the `estimate_total_trees()` function, which propagates uncertainty from the previously fitted alpha model into posterior estimates of total abundance. The model uses weakly informative priors on total abundance and leverages the fitted size–abundance structure to infer the number of trees consistent with the observed size distribution.


### Estimating total abundance across sites

```{r estimate-abundance-loop, eval = FALSE}
# Storage objects
# Initialize storage
results_abundance_model <- list()
results_abundance_summary <- data.frame()


abundance_runtime <- data.frame(
siteID = character(),
elapsed_sec = numeric(),
stringsAsFactors = FALSE
)

for (site in names(results_model)) {


message(paste("Estimating total abundance for site:", site))


# Extract fitted alpha model for this site
alpha_fit <- results_model[[site]]


# Time the abundance estimation
t_start <- Sys.time()


trees <- estimate_total_trees(
alpha_model_output = alpha_fit,
N_tot_prior_mean = 1250,
N_tot_prior_sd = 625,
warmup = 6000,
iter = 9000,
siteID = site
)


t_end <- Sys.time()


# Store full model output
results_abundance_model[[site]] <- trees


# Store posterior summary for N_tot
if (!is.null(trees$posterior_summary)) {
results_abundance_summary <- rbind(
results_abundance_summary,
trees$posterior_summary
)
}


# Store runtime
abundance_runtime <- rbind(
abundance_runtime,
data.frame(
siteID = site,
elapsed_sec = as.numeric(difftime(t_end, t_start, units = "secs"))
)
)
}


# Per-site runtimes
abundance_runtime


# Summary statistics
abundance_runtime_summary <- abundance_runtime %>%
summarise(
n_sites = n(),
mean_sec = mean(elapsed_sec),
median_sec = median(elapsed_sec),
sd_sec = sd(elapsed_sec),
min_sec = min(elapsed_sec),
max_sec = max(elapsed_sec)
)


abundance_runtime_summary

```

### Interpretation


The resulting posterior distributions of N_tot represent site-level estimates of total stem abundance, conditional on the inferred size–abundance scaling and prior information on plausible stand densities. These estimates can be used directly for site-to-site comparisons, downstream modeling, or integration with forest demographic and carbon accounting workflows.

---

## Visualizing Abundance Estimates

We can quickly compare estimated abundance values to previously reported values for validation. The 1:1 line in red would represent a "perfect" replication of measured data by our interpolation:

```{r}

abundancedata<-full_neon_data%>%
  rename(dbh=stemDiameter)%>%
  filter(dbh>=10 & dbh<=50)%>%
  group_by(siteID)%>%
  summarize(abundance = n())%>%
  data.frame()%>%
  rename(site = siteID)

ggplot(results_abundance_summary%>%
         inner_join(abundancedata), aes(x = abundance, y = mean, label = site)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(nudge_y = 0.1) +
  labs(x = "Reported Abundance", y = "ScalingFromSky Estimated Abundance") +
  theme_minimal() +
  geom_smooth(method = "lm")
```


