---
title: "ForestForTrees: Estimating Forest Size-Abundance Parameters"
author: "Adam Eichenwald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ForestForTrees: Estimating Forest Size-Abundance Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

Understanding how trees are distributed by size within forests is fundamental to ecology, management, and conservation. Tree size distributions shape forest structure, carbon storage, biodiversity, and resilience. Across many forest types, smaller trees are far more numerous than larger trees, and mid-sized trees often follow a predictable **power-law (Pareto) distribution**, where abundance declines systematically with size.

Remote sensing offers a powerful way to capture forest structure over broad areas. **LiDAR** provides detailed vertical information but is often limited in coverage, while high-resolution **RGB imagery** is more widely available but primarily captures canopy trees. Because many mid-sized and understory trees may be occluded, simply counting visible trees can underestimate true forest structure.

The **ForestForTrees R package** leverages the known form of the size-abundance distribution to infer the obscured portion of the forest. By fitting a Pareto distribution to the observable portion of the data, the package interpolates the expected parameters for the **10–50 cm DBH range**, allowing estimation of both the **Pareto exponent (α)** and the **total number of trees (N_tot)**, including those potentially beneath the canopy and invisible in a remotely sensed image.

The workflow requires pre-processed tree-level data (DBH) and optionally site-specific covariates such as **leaf area index (LAI)**. It focuses on trees in the 10–50 cm DBH range, where the Pareto distribution generally holds, while smaller (<10 cm) and larger (>50 cm) trees naturally deviate from this pattern for ecological reasons, independent of remote sensing visibility (Eichenwald et al., 2025).

The workflow includes:

1. Identifying the observable range of tree sizes that follow a power-law distribution.
2. Estimating the Pareto exponent (α) for the 10–50 cm DBH range.
3. Estimating the total number of trees (N_tot) for the area of interest, including those potentially occluded.
4. Generating spatially explicit maps of forest structural parameters across grid cells.

This vignette demonstrates a complete workflow using sample data from the Harvard Forest ForestGEO project. A separate vignette is included that shows the same workflow on NEON data with a benchmark. Although in our example the dataset was derived from canopy height rasters and segmentation of tree crowns, the methods in this package work with vector data (tree-level attributes such as DBH and height). The package itself does not process raw rasters; instead, it assumes a pre-processed dataset of individual crowns or trees.

## Load Packages and Data

We begin by loading the necessary libraries, including `ForestForTrees`, and our sample dataset.

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(VGAM)
library(ForestForTrees)

# Load the sample data included with the package
data("harv_data_sample")
```

## Step 1: Prepare Tree Data

ForestForTrees does not perform crown segmentation or DBH estimation itself; instead, these data are typically generated in earlier steps using other tools. For example, crown polygons can be segmented from LiDAR point clouds or high-resolution RGB imagery using packages such as lidR in R (Roussel et al., 2020) or DeepForest in Python (Weinstein et al., 2020), respectively. Geospatial packages such as sf in R can then import and analyze crown shapefiles, while allometric functions in itcSegment can be used to convert crown dimensions (e.g., height, area, perimeter) into DBH. In addition, site-level Leaf Area Index (LAI) is required to adjust estimates for potential segmentation errors. LAI can be obtained from field measurements or from remote-sensing products such as MODIS LAI or GEDI Plant Area Index (PAI).

Users can obtain crown segmentation data from Weinstein et al. (2024), available via Zenodo at https://doi.org/10.5281/zenodo.10926344, or from Weinstein et al. (2022) at https://doi.org/10.5281/zenodo.3459802. These datasets include delineated tree crowns and associated tree height estimates, making them suitable for calculating allometrically derived DBH values.

Users can also practice on the sample shapefile `harvardshapefile` included in this package. The shapefile represents a larger section of the Harvard Forest ForestGEO site and includes UTM-referenced crown boundaries that can be used for demonstration purposes.

```{r eval = FALSE}
library(itcSegment)

# Suppose `harvardshapefile` is your shapefile of tree crowns
# Make sure it has columns for height and geometry
harvardshapefile$perimeter <- as.numeric(sf::st_perimeter(harvardshapefile))
harvardshapefile$area <- as.numeric(sf::st_area(harvardshapefile))

# Estimate crown diameter from perimeter and area
harv_data_sample <- harvardshapefile %>%
  mutate(
    Diameter = 0.5 * sqrt(perimeter^2 - (8 * area)),
    # Calculate DBH using formula from Jucker et al. (2017)
    dbh = dbh(H = Max_Height, CA = Diameter, biome = 0)
  ) %>%
  data.frame()
```

Once we have estimates of DBH for each segmented tree, we can begin the process of working with ForestForTrees. Here we load our pre-processed tree data, `harv_data_sample`, which contains attributes such as `IDhectbest` (a unique identifier for each hectare plot) and `dbh` (diameter at breast height).

```{r}
# Filter for a specific plot for a single-site example
df_tile <- harv_data_sample %>%
  filter(IDhectbest == 1, !is.na(dbh)) %>%
  select(dbh)
```

## Step 2: Determine Observation Window

The size-abundance distribution of trees often follows a power-law, but for data from remote sensing part of this distribution is hidden due to smaller trees being overshadowed by larger trees in the canopy. We use a kernel density estimation (KDE) and bootstrapping approach to identify the potential breakpoint at which the observable data deviates from the predicted distribution. The `potential_break` function determines the point ($x_{min}$) where the observation window begins.

The function returns a named list with the following components:

- `potential_breakpoint`: **numeric scalar** giving the estimated lower log10 size bound for consistent detection.
- `bootstrap_kde_log`: **data frame** containing the log-scaled KDE curve (`log_x`, `mean_log_density`) used to identify peaks.
- `original_data_trimmed`: **numeric vector** of the DBH values after trimming by `trim_max`.
- `original_raw_data_df`: **data frame** containing the original, unmodified input data.
- `trim_max_value`: **numeric scalar** recording the `trim_max` value used in the function.

```{r}
kde_output <- potential_break(df_tile)
```

We can inspect the output to see the estimated breakpoint.

```{r results='asis'}
cat("### Estimated Breakpoint\n")
cat(10^kde_output$potential_breakpoint, "cm")
```

This is only a first pass at the breakpoint. To refine it, we apply `truncate_filter()`, which uses segmented regression on the bootstrapped KDE curve to finalize the lower breakpoint and identify the upper truncation point (generally 50 cm DBH). This truncation defines the observational window for downstream analysis.

The function returns a list with:

- `final_breakpoint`: **numeric scalar** giving the finalized lower log10 size bound.
- `bayesian_data`: **data frame** of the original tree data restricted to the observational window.
- `kerneldens_logtransform`: **data frame** of the full log-scaled KDE, useful for plotting.

```{r}
trunc_output <- truncate_filter(kde_output)
```

After running this function, `trunc_output$bayesian_data` contains the subset of trees within the consistent detection window and is used for fitting the truncated Pareto model in the next step.

## Step 3: Parameter Estimation

Next, we fit a Bayesian Pareto model to estimate the power-law exponent, α, using `fit_alpha_model()`. This step accounts for site-level LAI, which influences detectability of smaller trees in remote sensing data, and allows incorporation of prior information about α.

The main arguments are:

- `filtered_data`: The **list** returned by `truncate_filter()`, containing the filtered data and breakpoint.
- `LAI`: A **numeric value** for site-level leaf area index on a 1–10 scale.
- `prior_mean` and `prior_sd`: **Numeric values** specifying the prior for α. In our workflow, we estimated these using a random forest model trained on FIA data.
- `iter`, `warmup`, `chains`, `cores`: Standard Stan MCMC settings.

The function returns a **list** with:

- `posterior_summary`: A **data frame** summarizing the posterior distribution of α, including R² for model fit.
- `stan_fit`: The **full Stan fit object**.

```{r message=FALSE, warning=FALSE}
fit <- fit_alpha_model(
  filtered_data = trunc_output,
  LAI = 5.426,
  prior_mean = 1.4,
  prior_sd = 0.3
)
```

Let's create a plot showing the resulting fitted curve:

```{r fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
dbh_seq <- seq(10, 50, length.out = 100)
plot_data <- data.frame(
  dbh = dbh_seq,
  fit_n_dbh = dpareto(dbh_seq, shape = fit$posterior_summary$mean, scale = 10)
)

ggplot(trunc_output$bayesian_data, aes(x = dbh)) +
  geom_line(data = plot_data, aes(x = dbh, y = fit_n_dbh), color = "red", linewidth = 1) +
  labs(title = "Estimated Size-Abundance (Density) Distribution",
       x = "Diameter at Breast Height (cm)",
       y = "Tree Density") +
  theme_bw()
```

Using the posterior samples of α from our fitted model, we estimate the total number of trees for the hectare plot. The priors for N_tot (mean = 1250, SD = 625) are derived from TreeMap (Riley et al. 2021), a tree-level imputation of FIA plot data across the conterminous US.

```{r}
trees <- estimate_total_trees(
  alpha_model_output  = fit,
  N_tot_prior_mean    = 1250,
  N_tot_prior_sd      = 625
)
```

## Step 4: Final Output

Visualize the results, where the y-axis density is converted to tree abundance.

```{r fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
dbh_seq <- seq(10, 50, length.out = 100)
plot_data <- data.frame(
  dbh = dbh_seq,
  fit_n_dbh = dpareto(dbh_seq, shape = fit$posterior_summary$mean, scale = 10) * trees$posterior_summary$mean
)

ggplot(trunc_output$bayesian_data, aes(x = dbh)) +
  geom_line(data = plot_data, aes(x = dbh, y = fit_n_dbh), color = "red", linewidth = 1) +
  labs(title = "Estimated Size-Abundance Distribution",
       x = "Diameter at Breast Height (cm)",
       y = "Number of Trees") +
  theme_bw()
```

## Step 5: Loop Over Plots

The entire workflow can be applied across multiple spatial units by iterating through them. The following loop applies the functions to all 1-hectare subplots within the Harvard Forest dataset.

```{r}
alpha_results <- list()
tree_results  <- list()

for (i in unique(harv_data_sample$IDhectbest)) {

  df_tile <- harv_data_sample %>%
    filter(IDhectbest == i, !is.na(dbh)) %>%
    select(dbh)

  if (nrow(df_tile) < 25) next

  kde_output <- tryCatch(potential_break(df_tile), error = function(e) NULL)
  if (is.null(kde_output)) next

  trunc_output <- tryCatch(truncate_filter(kde_output), error = function(e) NULL)
  if (is.null(trunc_output)) next

  fit <- tryCatch(
    fit_alpha_model(
      filtered_data = trunc_output,
      LAI           = 3.5,
      prior_mean    = 1.4,
      prior_sd      = 0.3
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) next

  trees <- tryCatch(
    estimate_total_trees(
      alpha_model_output  = fit,
      N_tot_prior_mean    = 1250,
      N_tot_prior_sd      = 625
    ),
    error = function(e) NULL
  )

  site_name <- paste0("HARV_", i)
  alpha_results[[site_name]] <- fit$posterior_summary
  if (!is.null(trees)) tree_results[[site_name]] <- trees$posterior_summary
}
```

## Extra Steps: Parallelization

For large sites with many grid cells, parallelization can substantially reduce runtime. The following code demonstrates this using `foreach` and `doParallel`.

```{r eval = FALSE}
library(foreach)
library(doParallel)

n_cores <- parallel::detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

alpha_tree_results <- foreach(
  i = unique(harv_data_sample$IDhectbest),
  .packages = c("dplyr", "ForestForTrees")
) %dopar% {

  df_tile <- harv_data_sample %>%
    filter(IDhectbest == i, !is.na(dbh)) %>%
    select(dbh)

  if (nrow(df_tile) < 25) return(NULL)

  kde_output   <- tryCatch(potential_break(df_tile), error = function(e) NULL)
  if (is.null(kde_output)) return(NULL)

  trunc_output <- tryCatch(truncate_filter(kde_output), error = function(e) NULL)
  if (is.null(trunc_output)) return(NULL)

  fit <- tryCatch(
    fit_alpha_model(
      filtered_data = trunc_output,
      LAI           = 3.5,
      prior_mean    = 1.4,
      prior_sd      = 0.3
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NULL)

  trees <- tryCatch(
    estimate_total_trees(
      alpha_model_output  = fit,
      N_tot_prior_mean    = 1250,
      N_tot_prior_sd      = 625
    ),
    error = function(e) NULL
  )

  list(alpha = fit$posterior_summary, trees = if (!is.null(trees)) trees$posterior_summary else NULL)
}

stopCluster(cl)
```

## Step 6: Gridded Mapping

Once estimates are available for all plots, we can generate spatially explicit maps of α and N_tot using `map_grid_estimates()`. This function joins the results data frame to a spatial polygon layer. Here, `harvardshapefile` is the crown segmentation sf object — the function builds a 1-hectare grid and assigns each cell the plot ID of the majority of crown centroids within it, matching the `IDhectbest` values in our results.

```{r}
library(sf)

# Collate alpha results into a data frame
alpha_df <- dplyr::bind_rows(alpha_results, .id = "site_name") %>%
  dplyr::mutate(IDhectbest = as.integer(sub("HARV_", "", site_name)))

trees_df <- dplyr::bind_rows(tree_results, .id = "site_name") %>%
  dplyr::mutate(IDhectbest = as.integer(sub("HARV_", "", site_name)))

# Join to spatial grid
data("harvardshapefile")

alpha_map <- map_grid_estimates(
  results_df     = alpha_df,
  results_id_col = "IDhectbest",
  crowns_sf      = harvardshapefile,
  crowns_id_col  = "IDhectbest"
)

trees_map <- map_grid_estimates(
  results_df     = trees_df,
  results_id_col = "IDhectbest",
  crowns_sf      = harvardshapefile,
  crowns_id_col  = "IDhectbest"
)
```

We can then plot the spatial outputs side by side:

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)

p_alpha <- ggplot(alpha_map) +
  geom_sf(aes(fill = mean), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(name = expression(alpha)) +
  theme_void() +
  labs(tag = "a")

p_ntot <- ggplot(trees_map) +
  geom_sf(aes(fill = mean), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(name = expression(N[tot]), option = "inferno") +
  theme_void() +
  labs(tag = "b")

p_alpha + p_ntot
```

The output polygons can also be exported as a shapefile or converted to raster format:

```{r eval = FALSE}
# Export as shapefile
sf::write_sf(alpha_map, "harv_alpha.shp")

# Convert to raster
library(terra)
r <- terra::rast(terra::vect(alpha_map), resolution = 0.001)
r_alpha <- terra::rasterize(terra::vect(alpha_map), r, field = "mean")
```

## Literature Cited

Eichenwald, A. J., Grady, J. M., Knott, J. A., Rodriguez, J. M., Weinstein, B. G., Orwig, D. A., & Record, S. (2025). Leveraging Remote Sensing and Theory to Predict Tree Size Abundance Distributions Across Space. *Global Ecology and Biogeography*, 34(8).

Riley, K. L., I. C. Grenfell, M. A. Finney, and J. M. Wiener. 2021. "TreeMap, a Tree-Level Model of Conterminous US Forests Circa 2014 Produced by Imputation of FIA Plot Data." *Scientific Data* 8: 11.

Weinstein, B. G., Marconi, S., Aubry-Kientz, M., Vincent, G., Senyondo, H., & White, E. P. (2020). DeepForest: A Python package for RGB deep learning tree crown delineation. *Methods in Ecology and Evolution*, 11(12), 1743–1751.

Weinstein, B. G., Marconi, S., Bohlman, S. A., Zare, A., Singh, A., Graves, S. J., & White, E. P. (2021). A remote sensing derived data set of 100 million individual tree crowns for the National Ecological Observatory Network. *eLife*, 10, e62922.

Ben Weinstein, Sergio Marconi, & Ethan White. (2022). Data for the NeonTreeEvaluation Benchmark (0.2.2) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.5914554

Weinstein, B. G., et al. (2024). Individual tree species maps for the National Ecological Observatory Network. *PLoS Biology*, 22(7).
